## NCAA Football Attendance 

Let's load the tidyverse.

```{r}
library(tidyverse)
```

And read in the attendance data:

```{r}
attendance <- read_csv("https://dwillis.github.io/sports-data-files/attendance.csv")
```

Using the tidyverse's `pivot_longer` function, turn the currently wide data into longer data and save it as a new dataframe called `attendance_longer`:

```{r}
attendance_longer <- attendance |> pivot_longer(cols = starts_with("20"), names_to = "year", values_to = "attendance")
```

Let's try to figure out whether football attendance has returned to "normal" by 2024, defining "normal" as 2019 figures. For that we'll need to isolate those two years using `filter()` and get rid of any NAs:

```{r}
attendance_two_years <- attendance_longer |> filter(year == 2019 | year == 2024) |> na.omit()
```

Now, in order to calculate a percentage change between those two years, we'll need to go back to wider data and again ditch teams where we don't have both years:

```{r}
attendance_wide <- attendance_two_years |> pivot_wider(names_from = year, values_from = attendance) |> na.omit()
```

Finally, we'll add a percentage change column to that dataframe (remember, it's (New-Old)/Old) and arrange it to see the results.

```{r}
attendance_wide <- attendance_wide |> mutate(pct_change = ((`2024` - `2019`)/`2019`)*100) |> arrange(pct_change)
```

Some of these clearly are influenced by performance on the field. Below, describe the most interesting results (positive or negative) and what you think might be going on with them. Feel free to do some googling and describe what other data would be helpful to clarify what the best stories here might be. What surprises you about these results?

**ANSWER**

The general trend in stadium building these days is to build smaller venues with a focus on higher-priced premium seating. How much is that impacting downward trends in attendance data. Northwestern feels like a school to focus on here which is taking a large loss in attendance from 2019 to 2024 as it undergoes stadium reconstruction. Northwestern is currently playing at a makeshift glorified high school field alongside Lake Michigan while its new stadium is built (expected to be done for the 2026 season). The new stadium will have fewer seats than its previous home venue at Ryan Field. Northwestern may see a year over year uptick in attendance relative to its temporary stadium, but still a decrease relative to 2019. 

How will projects like Northwestern impact future projects? And is the trend toward premium seating going to price out fans — again lowering attendance with fewer people sitting in/able to afford regular seats? This is compared to a school like Temple which also saw a massive decrease in attendance, but plays in an NFL stadium. Obviously Temple is never going to fill an NFL stadium, so are they even offering all the possible seats for sale? And is there a benefit to playing in a massive stadium, off campus nonetheless, versus building a smaller venue closer to campus that might be more accessible and thus have a better attendance percentage — albeit still with fewer people actually at the games? 
